### 2011-04-30 Document/LaTeX/Article.tex
Initial check-in.

%
% Complete documentation on the extended LaTeX markup used for Insight
% documentation is available in ``Documenting Insight'', which is part
% of the standard documentation for Insight.  It may be found online
% at:
%
%     http://www.itk.org/

\documentclass{InsightArticle}

\usepackage[dvips]{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}

%  This is a template for Papers to the Insight Journal. 
%  It is comparable to a technical report format.

% The title should be descriptive enough for people to be able to find
% the relevant document. 
\title{Boolean Operations for Surfaces in VTK Without External Libraries}

% 
% NOTE: This is the last number of the "handle" URL that 
% The Insight Journal assigns to your paper as part of the
% submission process. Please replace the number "1338" with
% the actual handle number that you get assigned.
%
\newcommand{\IJhandlerIDnumber}{1338}

% Increment the release number whenever significant changes are made.
% The author and/or editor can define 'significant' however they like.
\release{1.00}

% At minimum, give your name and an email address.  You can include a
% snail-mail address if you like.
\author{Cory Quammen and Russell M. Taylor II}
\authoraddress{Department of Computer Science\\
             The University of North Carolina at Chapel Hill
}

\begin{document}

%
% Add hyperlink to the web location and license of the paper.
% The argument of this command is the handler identifier given
% by the Insight Journal to this paper.
% 
\IJhandlefooter{\IJhandlerIDnumber}

\ifpdf
\else
   %
   % Commands for including Graphics when using latex
   % 
   \DeclareGraphicsExtensions{.eps,.jpg,.gif,.tiff,.bmp,.png}
   \DeclareGraphicsRule{.jpg}{eps}{.jpg.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.gif}{eps}{.gif.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.tiff}{eps}{.tiff.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.bmp}{eps}{.bmp.bb}{`convert #1 eps:-}
   \DeclareGraphicsRule{.png}{eps}{.png.bb}{`convert #1 eps:-}
\fi

\maketitle

\ifhtml
\chapter*{Front Matter\label{front}}
\fi

% The abstract should be a paragraph or two long, and describe the
% scope of the document.
\begin{abstract}
\noindent
A previous VTK Journal article provided a VTK class that enabled boolean operations on surfaces defined by \code{vtkPolyData} objects. While the capabilities introduced by the previous contribution satisfied numerous request on the vtk-developers and vtk-users email lists, the class could not be incorporated into VTK because of a licensing incompatibility with the GTS library used to perform the boolean operations. This article describes a set of classes than enable the same boolean operations as the previous contribution using only VTK classes. Our contribution also has some feature enhancements over the previous contribution, namely,  support for meshes with more than one connected component and interpolation of point data at points introduced by the splitting procedure.

\end{abstract}

\IJhandlenote{\IJhandlerIDnumber}

\tableofcontents

\section{Clipping One Surface with Another Surface}

The most complex part of the computing boolean operations on a surface mesh is splitting one input surface mesh by the other.

\subsection{Identifying Surface-Surface Intersection}

Identifying the intersection of a mesh with another mesh involves identifying the intersection between each triangle in the first mesh with the triangles that intersect it in the second mesh. A triangle-triangle intersection test yields two end points of the line segment defining the intersection for non-coplanar triangles; this contribution does not explicitly handle intersections between two coplanar triangles. The points from these intersections are merged using a \code{vtkMergePoints} object, leading to many fewer connected components than if the line segments were stored as ``line soup''.

\subsection{Splitting the Mesh}

\section{Mesh-to-Mesh Signed Distance Calculation}

\section{Boolean Operations}

\subsection{Union}

\subsection{Intersection}

\subsection{Difference}

\section{Limitations}

These classes do not properly handle intersections between coplanar triangles. When such an intersection event has been detected, the intersection information is discarded. Such intersections likely occur when portions of the meshes are immediately adjacent to each other, however, and their exclusion from the surface-to-surface intersection information will usually not affect the mesh splitting results.

\section{Software Requirements}

You need to have the following software installed:

% The {itemize} environment uses a bullet for each \item.  If you want the 
% \item's numbered, use the {enumerate} environment instead.
\begin{itemize}
  \item  Insight Toolkit 2.4.
  \item  CMake 2.2
\end{itemize}

Note that other versions of the Insight Toolkit are also available in the
testing framework of the Insight Journal. Please refere to the following page
for details

\url{http://www.insightsoftwareconsortium.org/wiki/index.php/IJ-Testing-Environment}

% The preceding sections will have been written in a gentler,
% introductory style.  You may also wish to include a reference
% section, documenting all the functions/exceptions/constants.
% Often, these will be placed in separate files and input like this:

\appendix

\section{This is an Appendix}

To create an appendix in a Insight HOWTO document, use markup like
this:

\begin{verbatim}
\appendix

\section{This is an Appendix}

To create an appendix in a Insight HOWTO document, ....

\section{This is another}

Just add another \section{}, but don't say \appendix again.
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Example on how to insert a figure
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
\center
%\includegraphics[width=0.8\textwidth]{RegistrationComponentsDiagram.eps}
\itkcaption[Registration Framework Components]{The basic components of the
registration framework are two input images, a transform, a metric, an
interpolator and an optimizer.}
\label{fig:RegistrationComponents}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Insert the bibliography using BibTeX
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plain}
\bibliography{Article}

\end{document}

### 2011-04-30 Document/LaTeX/Article.tex
Added text on intersection identification.

@@
 A triangle-triangle intersection test yields two end points of the line segment defining the intersection for non-coplanar triangles; this contribution does not explicitly handle intersections between two coplanar triangles.
-The points from these intersections are merged using a \code{vtkMergePoints} object, leading to many fewer connected components than if the line segments were stored as ``line soup''.
+
+Oriented bounding box (OBB) trees are used to accelerate the identification of triangle-triangle intersections between meshes.
+Two \code{vtkOBBTree} objects are instantiated, one for each input mesh.
+The method \code{vtkOBBTree::IntersectWithOBBTree()} is then called on one OBB tree with the second OBB tree and a callback function passed as parameters.
+The callback function performs exact triangle-triangle intersection tests between the triangles in overlapping nodes from the two OBB trees and stores the results in several data structures used later in the algorithm.
+
+End points from the triangle-triangle intersections are stored in a \code{vtkPoints} object and the line cells are stored in a \code{vtkCellArray}.
+The points  are merged using a \code{vtkMergePoints} object, leading to many fewer connected components than if the line segments were stored as ``line soup''.
+In addition, each end point is tested to see if it lies on each edge of the two input triangles.
+If so, the index of the triangle, the index of the edge, and the index of the intersecting line is associated with the index of the endpoint in a \code{std::multimap}.
+
+The end point-edge intersection information is used to split the fully connected intersection lines to respect the topology of the input mesh.
+Because the end points of the intersection lines are merged during construction, they need to be split where the mesh is split.
+The end result should be that a copy of each line end point is produced for each set of triangles that share an edge.
+This can be achieved by iterating over the elements in the end point-edge intersection \code{std::multimap}, creating a copy of the point, and removing all records in the multimap where the cell ID is the same as one identified as being a neighbor of the current cell being processed connected at the edge in question.
 

### 2011-04-30 Document/LaTeX/Article.tex
Added text on mesh splitting.

@@
 
-The end point-edge intersection information is used to split the fully connected intersection lines to respect the topology of the input mesh.
+The end point-edge intersection information is used to split the fully connected intersection lines to respect the topology of each input mesh.
 Because the end points of the intersection lines are merged during construction, they need to be split where the mesh is split.
@@
 \subsection{Splitting the Mesh}
+
+After the intersection lines are split for a mesh, the next step is to identify candidate cells for splitting. A cell is a candidate cell in two cases:
+
+\begin{enumerate}
+\item The cell index is in the intersection line map, meaning that intersection lines lie inside the cell.
+\item The cell is the neighbor of a cell identified by case 1.
+\end{enumerate}
+
+The second case is important because one of the mesh-mesh intersection lines may have an endpoint on the edge of one cell, but no line that uses that endpoint on the cell neighbor across that edge.
+The cell neighbor needs to be split to avoid introducing a T-junction and therefore a hole in the output mesh.
+
+Splitting proceeds on a cell-by-cell basis. For each cell that needs splitting, the following data is gathered:
+
+\begin{enumerate}
+\item The three points that define the cell.
+\item The boundary lines of the cell.
+\item The split intersection lines that lie in the cell.
+\item The points that define the lines in 3.
+\item Points from neighboring cells that lie on a cell edge but which were not already added in 4.
+\end{enumerate}
+
+The lines collected above are used to constrain the new triangulation of the points.
+That is to say, these lines should be present in the output mesh.
+Special processing of the boundary lines is required to achieve the desired split.
+Specifically, boundary lines must be split at points that lie on them.
+
+The \code{vtkDelaunay2D} class can be used to achieve a constrained triangulation.
+A transformation of the points collected above that rotates them to the XY-plane is calculated is computed and set as the transform for the \code{vtkDelaunay2D} object.
+A new \code{vtkPolyData} object that contains the collected points and lines is defined and passed as both the input and constraint source to a \code{vtkDelaunay2D} object.
+The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.
 

### 2011-05-10 Document/LaTeX/Article.tex
Added more text.

@@
 
+\def\x{{\mathbf x}}
+\def\m{{\mathbf m}}
+\def\v{{\overrightarrow{\mathbf v}}}
+
 %  This is a template for Papers to the Insight Journal. 
@@
 % snail-mail address if you like.
-\author{Cory Quammen and Russell M. Taylor II}
+\author{Cory Quammen, Chris Weigle, and Russell M. Taylor II}
 \authoraddress{Department of Computer Science\\
@@
 This article describes a set of classes than enable the same boolean operations as the previous contribution using only VTK classes.
-Our contribution also has some feature enhancements over the previous contribution, namely,  support for meshes with more than one connected component and interpolation of point data at points introduced by the splitting procedure.
+Our contribution also has some feature enhancements over the previous contribution, namely, preservation of mesh topology as well as passing point data and cell data through to outputs.
 
@@
 
+\section{Introduction}
+
+Let $A$ and $B$ be sets. A third set $C$ can be defined from these sets via a \emph{boolean operation}. Three common boolean operations on sets are:
+
+\begin{description}
+
+\item Union
+
+ $C = A \cup B = \{ x \, | \, x \in A \text{ or } x \in B \}$
+
+\item Intersection
+
+$C = A \cap B = \{ x \, | \, x \in A \text{ and } x \in B \}$
+
+\item Difference
+
+$C = A - B = \{ x \, | \, x \in A \text{ and } x \notin B \}$
+
+\end{description}
+
+In geometric modeling, a closed surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$.
+Then given a second surface mesh $M_B$, a boolean operation may be applied to obtain a third surface mesh $M_C$.
+These simple operations can be used to define complex geometries.
+
+Rather than convert two mesh representations to a set of points, perform the boolean operations on those points, and reconstruct a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
+This article describes an implementation of those methods in VTK.
+
+\section{Mesh-to-Mesh Signed Distance Calculation}
+
+The distance field $f(\x)$ of a mesh is defined as the distance from $\x$ to the nearest point on the mesh.
+The signed distance field is similar to the distance field, but the sign of the field is negative for locations $\x$ inside the mesh and positive for locations $\x$ outside the mesh.
+To determine whether $\x$ is inside or outside the mesh, a vector $\v = \x - \m$ is defined where $\m$ is the nearest point on the mesh.
+B\ae rentzen and Aan\ae s proved that the sign of the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ is the same as the sign of the distance field for closed, orientable 2-manifold surfaces in 3d Euclidean space \cite{Baerentzen2005}.
+
+The angle-weighted pseudo-normal is defined separately for faces, edges, and points in a mesh.
+For faces, it is simply the normal of the triangle.
+For an edge, it is the average of the faces that share the edge.
+For a point, it is the average of the angle-weighted normals from each face defined by the point where the weight is the angle between the two face edges incident to the point.
+
+We introduce the \code{vtkImplicitPolyData} class that defines an implicit signed distance function given an input \code{vtkPolyData}.
+This class is a subclass of \code{vtkImplicitFunction}, so it can be used by classes that operate on \code{vtkImplicitFunction} objects (see, e.g., \code{vtkSampleFunction}).
+The function value is the signed distance, and its gradient value is the angle-weighted pseudo-normal.
+
+To support boolean operations, we introduce another class, \code{vtkPolyDataDistance}, that computes the distance from the first input \code{vtkPolyData} to a second.
+Optionally, the distance field from the second input \code{vtkPolyData} to the first can also be computed.
+These distances are computed at point locations and are stored as point data.
+
+\section{Boolean Operations Using Signed Distance}
+
+Boolean operations can be computed with only the signed distance field for each mesh.
+As the sign of the distance field at a point in the mesh corresponds to whether that point is inside (negative) or outside (positive) the other mesh, the surfaces bounding the volumes produced by the boolean operations are:
+
+\begin{description}
+
+\item Union
+
+The union of two meshes is the set of cells in each mesh such that the distance from each cell point to the other mesh is greater than or equal to zero.
+
+\item Intersection
+
+The intersection of two meshes is the set of cells in each mesh such that the distance from each cell point to the other mesh is less than or equal to zero.
+
+\item Difference
+
+Given meshes $A$ and $B$, the difference $A-B$ is defined by the cells of $A$ with positive distance from $B$ and the cells of $B$ with negative distance from $A$.
+
+\end{description}
+
 \section{Clipping One Surface with Another Surface}
 
-The most complex part of the computing boolean operations on a surface mesh is splitting one input surface mesh by the other.
+While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the union, intersection, and difference surfaces, there is a problem.
+\code{vtkClipPolyData} assumes that the implicit function is piecewise linear within each cell of the \code{vtkPolyData} being clipped.
+However, in general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field.
+The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the zero-level of the distance field.
+Consequently, the clipped portions of each input geometry will not match at the zero-level where they should fuse seemlessly [INSERT FIGURE].
 
-\subsection{Identifying Surface-Surface Intersection}
+Instead of directly clipping the input meshes based on the signed distance evaluated at mesh points, it is necessary to split each input mesh at the actual intersection of the two surfaces (hereafter referred to as the \emph{surface intersection}).
+This step is the most complex part of computing boolean operations on surface meshes.
 
-Identifying the intersection of a mesh with another mesh involves identifying the intersection between each triangle in the first mesh with the triangles that intersect it in the second mesh.
+\subsection{Identifying the Surface Intersection}
+
+Identifying the surface intersection of a mesh with another mesh involves identifying the intersection between each triangle in the first mesh with the triangles that intersect it in the second mesh.
 A triangle-triangle intersection test yields two end points of the line segment defining the intersection for non-coplanar triangles; this contribution does not explicitly handle intersections between two coplanar triangles.
@@
 End points from the triangle-triangle intersections are stored in a \code{vtkPoints} object and the line cells are stored in a \code{vtkCellArray}.
-The points  are merged using a \code{vtkMergePoints} object, leading to many fewer connected components than if the line segments were stored as ``line soup''.
+The points  are merged using a \code{vtkPointLocator} object with a small tolerance, leading to many fewer connected components than if the line segments were stored as ``line soup''.
 In addition, each end point is tested to see if it lies on each edge of the two input triangles.
-If so, the index of the triangle, the index of the edge, and the index of the intersecting line is associated with the index of the endpoint in a \code{std::multimap}.
+If so, the index of the triangle, the index of the edge, and the index of the intersecting line is associated with the index of the endpoint in a \code{std::multimap} for later use.
 
@@
 The end result should be that a copy of each line end point is produced for each set of triangles that share an edge.
-This can be achieved by iterating over the elements in the end point-edge intersection \code{std::multimap}, creating a copy of the point, and removing all records in the multimap where the cell ID is the same as one identified as being a neighbor of the current cell being processed connected at the edge in question.
+This can be achieved by iterating over the elements in the end point-edge intersection \code{std::multimap}, creating a copy of the point, and removing all records in the multimap where the cell ID is the same as one identified as sharing the the edge in question.
 
@@
 The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.
-
-\section{Mesh-to-Mesh Signed Distance Calculation}
-
-\section{Boolean Operations}
-
-\subsection{Union}
-
-\subsection{Intersection}
-
-\subsection{Difference}
 

### 2011-05-10 Document/LaTeX/Article.tex
Edited through section 2.

@@
 \noindent
-A previous VTK Journal article provided a VTK class that enabled boolean operations on surfaces defined by \code{vtkPolyData} objects.
-While the capabilities introduced by the previous contribution satisfied numerous request on the vtk-developers and vtk-users email lists, the class could not be incorporated into VTK because of a licensing incompatibility with the GTS library used to perform the boolean operations.
-This article describes a set of classes than enable the same boolean operations as the previous contribution using only VTK classes.
-Our contribution also has some feature enhancements over the previous contribution, namely, preservation of mesh topology as well as passing point data and cell data through to outputs.
+We have written a set of classes than enable boolean operations on surface meshes defined in \code{vtkPolyData} objects using only VTK classes.
+Our contribution offers some feature enhancements over previous solutions, namely, preservation of mesh topology and passing point data and cell data through to outputs.
 
@@
 
-Let $A$ and $B$ be sets. A third set $C$ can be defined from these sets via a \emph{boolean operation}. Three common boolean operations on sets are:
+Let $A$ and $B$ be sets. A \emph{boolean operation} can be used to define a third set $C$. Three common boolean operations on sets are:
 
@@
 
-In geometric modeling, a closed surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$.
-Then given a second surface mesh $M_B$, a boolean operation may be applied to obtain a third surface mesh $M_C$.
+In geometric modeling, a surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$ if $M_A$ is a closed, orientable 2-manifold surface in 3D.
+Given a second surface mesh $M_B$ with the same properties, a boolean operation may be applied to obtain a third surface mesh $M_C$.
 These simple operations can be used to define complex geometries.
@@
 Rather than convert two mesh representations to a set of points, perform the boolean operations on those points, and reconstruct a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
-This article describes an implementation of those methods in VTK.
+This article describes an implementation of those methods to realize boolean operations on surfaces in VTK.
 
@@
 To determine whether $\x$ is inside or outside the mesh, a vector $\v = \x - \m$ is defined where $\m$ is the nearest point on the mesh.
-B\ae rentzen and Aan\ae s proved that the sign of the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ is the same as the sign of the distance field for closed, orientable 2-manifold surfaces in 3d Euclidean space \cite{Baerentzen2005}.
+The sign of the distance field is the same as the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ when the mesh is a closed, orientable 2-manifold surfaces in 3D Euclidean space \cite{Baerentzen2005}.
 
@@
 For faces, it is simply the normal of the triangle.
-For an edge, it is the average of the faces that share the edge.
-For a point, it is the average of the angle-weighted normals from each face defined by the point where the weight is the angle between the two face edges incident to the point.
+For an edge, it is the average of the normals for the faces that share the edge.
+For a point, it is the sum of the angle-weighted normals from each face where the weight for a face normal is the angle between the two edges of that face incident to the point.
 
-We introduce the \code{vtkImplicitPolyData} class that defines an implicit signed distance function given an input \code{vtkPolyData}.
+We introduce the \code{vtkImplicitPolyData} class that defines a signed distance function given an input \code{vtkPolyData}.
 This class is a subclass of \code{vtkImplicitFunction}, so it can be used by classes that operate on \code{vtkImplicitFunction} objects (see, e.g., \code{vtkSampleFunction}).
-The function value is the signed distance, and its gradient value is the angle-weighted pseudo-normal.
+The method XXXX is overridden to return the signed distance, and XXXX is overridden to return the angle-weighted pseudo-normal.
 
@@
 Optionally, the distance field from the second input \code{vtkPolyData} to the first can also be computed.
-These distances are computed at point locations and are stored as point data.
+These distances are computed at points in the mesh and are stored as a point data field named ``Distance''.
 
@@
 \item The boundary lines of the cell.
-\item The split intersection lines that lie in the cell.
+\item The split intersection lines that lie in the cell (if any).
 \item The points that define the lines in 3.
@@
 Specifically, boundary lines must be split at points that lie on them.
+Boundary line splitting is achieved by sorting points on the boundary lines according to the angle between the first edge of the cell and the vector formed by subtracting the cell center from the point and connecting points adjacent to each other with lines.
+Because the last boundary point is connected to the first, the absolute order of the boundary points is not important.
 
-The \code{vtkDelaunay2D} class can be used to achieve a constrained triangulation.
+The \code{vtkDelaunay2D} class can be used to determine a triangulation from the cell, surface intersection lines, and boundary points and lines.
 A transformation of the points collected above that rotates them to the XY-plane is calculated is computed and set as the transform for the \code{vtkDelaunay2D} object.
@@
 The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.
+If a point in the surface intersection is within some small tolerance distance from a point in the surface mesh, the point in the surface mesh is used instead.
 
@@
 
-You need to have the following software installed:
-
-% The {itemize} environment uses a bullet for each \item.  If you want the 
-% \item's numbered, use the {enumerate} environment instead.
-\begin{itemize}
-  \item  Insight Toolkit 2.4.
-  \item  CMake 2.2
-\end{itemize}
-
-Note that other versions of the Insight Toolkit are also available in the
-testing framework of the Insight Journal. Please refere to the following page
-for details
-
-\url{http://www.insightsoftwareconsortium.org/wiki/index.php/IJ-Testing-Environment}
-
-% The preceding sections will have been written in a gentler,
-% introductory style.  You may also wish to include a reference
-% section, documenting all the functions/exceptions/constants.
-% Often, these will be placed in separate files and input like this:
-
-\appendix
-
-\section{This is an Appendix}
-
-To create an appendix in a Insight HOWTO document, use markup like
-this:
-
-\begin{verbatim}
-\appendix
-
-\section{This is an Appendix}
-
-To create an appendix in a Insight HOWTO document, ....
-
-\section{This is another}
-
-Just add another \section{}, but don't say \appendix again.
-\end{verbatim}
+The source code contributed with this article has been build against VTK 5.8.
 

### 2011-05-10 Document/LaTeX/Article.tex
Round of edits from Russ.

@@
 \noindent
-We have written a set of classes than enable boolean operations on surface meshes defined in \code{vtkPolyData} objects using only VTK classes.
-Our contribution offers some feature enhancements over previous solutions, namely, preservation of mesh topology and passing point data and cell data through to outputs.
+We have written a set of classes than enable boolean operations on the space contained within surface meshes defined in \code{vtkPolyData} objects using only VTK classes.
+In addition to being compatible with the VTK license, our contribution preserves mesh topology and passes point data and cell data through to outputs.
 
@@
 \section{Introduction}
+\label{sec:Introduction}
 
@@
 
-\item Union
+\item union
 
@@
 
-\item Intersection
+\item intersection
 
@@
 
-\item Difference
+\item difference
 
@@
 
-In geometric modeling, a surface mesh $M_A$ may be considered the boundary of a space-filling set of 3D points $A$ if $M_A$ is a closed, orientable 2-manifold surface in 3D.
+In geometric modeling, a closed, orientable 2-manifold surface mesh in 3D $M_A$ may be considered the boundary of an infinite set of 3D points $A$.
 Given a second surface mesh $M_B$ with the same properties, a boolean operation may be applied to obtain a third surface mesh $M_C$.
@@
 
-Rather than convert two mesh representations to a set of points, perform the boolean operations on those points, and reconstruct a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
+Rather than converting two mesh representations to a set of points, performing the boolean operations on those points, and reconstructing a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
 This article describes an implementation of those methods to realize boolean operations on surfaces in VTK.
@@
 
-The distance field $f(\x)$ of a mesh is defined as the distance from $\x$ to the nearest point on the mesh.
-The signed distance field is similar to the distance field, but the sign of the field is negative for locations $\x$ inside the mesh and positive for locations $\x$ outside the mesh.
+The distance field $f(\x)$ of a mesh is defined as the distance from $\x$ to the nearest point on the surface defined by the mesh.
+The signed distance field is similar to the distance field, but the sign of the field is negative for locations $\x$ inside the space bounded by the mesh and positive for outside locations.
 To determine whether $\x$ is inside or outside the mesh, a vector $\v = \x - \m$ is defined where $\m$ is the nearest point on the mesh.
-The sign of the distance field is the same as the dot product of $\v$ with the \emph{angle-weighted pseudo-normal} of the mesh at $\m$ when the mesh is a closed, orientable 2-manifold surfaces in 3D Euclidean space \cite{Baerentzen2005}.
+The sign of the distance field is the same as the dot product of $\v$ with the \emph{angle-weighted pseudonormal} of the mesh at $\m$ when the mesh is a closed, orientable 2-manifold surfaces in 3D Euclidean space \cite{Baerentzen2005}.
 
-The angle-weighted pseudo-normal is defined separately for faces, edges, and points in a mesh.
-For faces, it is simply the normal of the triangle.
+The angle-weighted pseudonormal is defined separately for faces, edges, and points in a mesh, and forms a discontinuous vector field over the surface defined by the mesh.
+For a face, it is simply the face normal.
 For an edge, it is the average of the normals for the faces that share the edge.
@@
 This class is a subclass of \code{vtkImplicitFunction}, so it can be used by classes that operate on \code{vtkImplicitFunction} objects (see, e.g., \code{vtkSampleFunction}).
-The method XXXX is overridden to return the signed distance, and XXXX is overridden to return the angle-weighted pseudo-normal.
+The method \code{double EvaluateFunction(double x[3])} is overridden to return the signed distance, and \code{void EvaluateGradient(double x[3], double g[3])} is overridden to return the angle-weighted pseudonormal.
 
@@
 
-\section{Boolean Operations Using Signed Distance}
+\section{Boolean Operations on Surfaces Using Signed Distance}
 
 Boolean operations can be computed with only the signed distance field for each mesh.
-As the sign of the distance field at a point in the mesh corresponds to whether that point is inside (negative) or outside (positive) the other mesh, the surfaces bounding the volumes produced by the boolean operations are:
+The sign of the distance field at a point in the mesh corresponds to whether that point is inside (negative), outside (positive), or on (zero) the other mesh; the surfaces bounding the volumes produced by the boolean operations described in Section \ref{sec:Introduction} can thus be defined as:
 
@@
 
-\item Union
+\item union
 
-The union of two meshes is the set of cells in each mesh such that the distance from each cell point to the other mesh is greater than or equal to zero.
+The set of cells in each mesh such that the distance from each cell point to the other mesh is greater than or equal to zero.
 
-\item Intersection
+\item intersection
 
-The intersection of two meshes is the set of cells in each mesh such that the distance from each cell point to the other mesh is less than or equal to zero.
+The set of cells in each mesh such that the distance from each cell point to the other mesh is less than or equal to zero.
 
-\item Difference
+\item difference
 
-Given meshes $A$ and $B$, the difference $A-B$ is defined by the cells of $A$ with positive distance from $B$ and the cells of $B$ with negative distance from $A$.
+The difference is defined as the set of cells of $M_A$ whose points are a non-negative distance from $M_B$ combined with the cells of $M_B$ whose points are a non-positive distance from $M_A$.
 
@@
 
-\section{Clipping One Surface with Another Surface}
+\section{Clipping One Surface Mesh with Another}
 
-While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the union, intersection, and difference surfaces, there is a problem.
-\code{vtkClipPolyData} assumes that the implicit function is piecewise linear within each cell of the \code{vtkPolyData} being clipped.
-However, in general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field.
-The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the zero-level of the distance field.
-Consequently, the clipped portions of each input geometry will not match at the zero-level where they should fuse seemlessly [INSERT FIGURE].
+While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, there is a problem.
+\code{vtkClipPolyData} treats the implicit function as piecewise linear within each cell of the \code{vtkPolyData} being clipped.
+In general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field.
+The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the true zero level of the distance field.
+Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seemlessly [INSERT FIGURE].
 
 Instead of directly clipping the input meshes based on the signed distance evaluated at mesh points, it is necessary to split each input mesh at the actual intersection of the two surfaces (hereafter referred to as the \emph{surface intersection}).
-This step is the most complex part of computing boolean operations on surface meshes.
+All points on the zero-level of the distance field are then completely defined by the points and edges of the surface intersection, and the surface intersection marks the accurate boundary between inside and outside portions of the mesh surface.
+We have defined a new class \code{vtkPolyDataIntersection} that computes the surface intersection between two surface meshes and can optionally split the input meshes by the surface intersection.
+Its operation is described below.
 
@@
 
-Identifying the surface intersection of a mesh with another mesh involves identifying the intersection between each triangle in the first mesh with the triangles that intersect it in the second mesh.
-A triangle-triangle intersection test yields two end points of the line segment defining the intersection for non-coplanar triangles; this contribution does not explicitly handle intersections between two coplanar triangles.
+Identifying the surface intersection between two surface meshes is the most complex part of the boolean operations algorithm.
+It involves identifying the intersection between each triangle in the first mesh with all triangles in the second mesh that intersect it.
+A triangle-triangle intersection test yields two endpoints (possibly the same) of the actual line segment defining the intersection when the triangles intersect.
+The intersection between overlapping and co-planar triangles is more complicated; \code{vtkPolyDataIntersection} does not explicitly handle intersections between two coplanar triangles.
 
@@
 The method \code{vtkOBBTree::IntersectWithOBBTree()} is then called on one OBB tree with the second OBB tree and a callback function passed as parameters.
-The callback function performs exact triangle-triangle intersection tests between the triangles in overlapping nodes from the two OBB trees and stores the results in several data structures used later in the algorithm.
+The callback function performs more accurate triangle-triangle intersection tests between the triangles in overlapping nodes from the two OBB trees and stores the results in several data structures used later in the algorithm.
 
@@
 The points  are merged using a \code{vtkPointLocator} object with a small tolerance, leading to many fewer connected components than if the line segments were stored as ``line soup''.
+Merging the end points also simplifies later operations.
 In addition, each end point is tested to see if it lies on each edge of the two input triangles.
-If so, the index of the triangle, the index of the edge, and the index of the intersecting line is associated with the index of the endpoint in a \code{std::multimap} for later use.
+If so, a key-value pair is stored in a \code{std::multimap} where the key is the ID of the endpoint and the value is a 3-tuple consisting of the index of the triangle, the index of the triangle edge, and the index of the intersecting line.
 
-The end point-edge intersection information is used to split the fully connected intersection lines to respect the topology of each input mesh.
-Because the end points of the intersection lines are merged during construction, they need to be split where the mesh is split.
-The end result should be that a copy of each line end point is produced for each set of triangles that share an edge.
-This can be achieved by iterating over the elements in the end point-edge intersection \code{std::multimap}, creating a copy of the point, and removing all records in the multimap where the cell ID is the same as one identified as sharing the the edge in question.
+The entries in the \code{std::multimap} described above are used to split the fully connected intersection lines to respect the topology of each input mesh.
+Because the end points of the intersection lines are merged during construction, they need to be split at locations where the mesh is split.
+For example, a mesh representing a cube often has duplicate points at the corners to store three different normals, one for each face incident to the point.
+These normals enable the appearance of sharp edges on the mesh.
+If the surface intersection on a cube splits an edge that connects two of these corners, then two points should be inserted at the split point rather than one to ensure that two face normals are defined at the split point.
+
+XXXX work on below paragraph again
+
+More generally, the end result of splitting the surface intersection should be that a copy of each line end point is produced for each set of triangles that share an edge.
+This can be achieved by iterating over the key-value pairs in the multimap, appending a copy of the point corresponding to the point ID in the key to the end of the \code{vtkPoints} object storing the endpoints, and removing all records in the multimap where the point ID is the same as the first point and the cell ID marks a cell that shares the edge identified in the first record.
+The point indices of the lines within the cells that share an edge are renumbered to the ID of the new point in the \code{vtkPoints} object.
+Finally, the point ID and the cell ID from one of the cells that shares the edge is stored in a point-to-cell map for future use in interpolating point data at the new point.
+It does not matter which cell ID is associated with a point on the edge of a cell because the interpolation results will be the same.
+
+The points in the input mesh and their associated point data are copied to the points in the output mesh.
+The points from the intersection lines are then appended to the point set in the output mesh, and the point data is interpolated at those points.
 
@@
 
-The lines collected above are used to constrain the new triangulation of the points.
-That is to say, these lines should be present in the output mesh.
+The lines collected above are used to constrain a new triangulation of the points.
+In other words, these lines will be present in the output mesh.
 Special processing of the boundary lines is required to achieve the desired split.
 Specifically, boundary lines must be split at points that lie on them.
-Boundary line splitting is achieved by sorting points on the boundary lines according to the angle between the first edge of the cell and the vector formed by subtracting the cell center from the point and connecting points adjacent to each other with lines.
+Boundary line splitting is achieved by sorting points on the boundary lines according to the angle between the first edge of the cell and the vector formed by subtracting the cell center from the point.
+Adjacent points in the sorted list are then connected with lines, include a line between the last and first points in the list.
 Because the last boundary point is connected to the first, the absolute order of the boundary points is not important.
@@
 The \code{vtkDelaunay2D} class can be used to determine a triangulation from the cell, surface intersection lines, and boundary points and lines.
-A transformation of the points collected above that rotates them to the XY-plane is calculated is computed and set as the transform for the \code{vtkDelaunay2D} object.
+A transformation of the points collected above that rotates them to the XY-plane is computed and set as the transform for the \code{vtkDelaunay2D} object.
 A new \code{vtkPolyData} object that contains the collected points and lines is defined and passed as both the input and constraint source to a \code{vtkDelaunay2D} object.
-The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map back to their original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.
+The line point indices are renumbered to point to their locations in the new \code{vtkPolyData} object, and a map from the original indices is created to remap the output cell indices from the triangulation to the point indices in the output mesh.
 If a point in the surface intersection is within some small tolerance distance from a point in the surface mesh, the point in the surface mesh is used instead.
+This prevents unintentional splitting of the surface mesh.
+
+\section{Results}
+
+XXXX Show pictures of various objects with boolean operations applied to them.
 
@@
 
-These classes do not properly handle intersections between coplanar triangles.
+The \code{vtkPolyDataIntersection} class operates only on triangulated surface meshes, and it does not properly handle intersections between coplanar triangles.
 When such an intersection event has been detected, the intersection information is discarded.
-Such intersections likely occur when portions of the meshes are immediately adjacent to each other, however, and their exclusion from the surface-to-surface intersection information will usually not affect the mesh splitting results.
+If this occurs, then the output of a boolean operation may not be exactly correct at the location of the co-planar triangles.
 
@@
 
-\begin{figure}
-\center
+%\begin{figure}
+%\center
 %\includegraphics[width=0.8\textwidth]{RegistrationComponentsDiagram.eps}
-\itkcaption[Registration Framework Components]{The basic components of the
-registration framework are two input images, a transform, a metric, an
-interpolator and an optimizer.}
-\label{fig:RegistrationComponents}
-\end{figure}
+%\itkcaption[Registration Framework Components]{The basic components of the
+%registration framework are two input images, a transform, a metric, an
+%interpolator and an optimizer.}
+%\label{fig:RegistrationComponents}
+%\end{figure}
 

### 2011-05-10 Document/LaTeX/Article.tex
Added figures.

@@
 The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the true zero level of the distance field.
-Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seemlessly [INSERT FIGURE].
+Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seamlessly.
+XXXX Figure
 
@@
 A triangle-triangle intersection test yields two endpoints (possibly the same) of the actual line segment defining the intersection when the triangles intersect.
+If the endpoints are the same, then the intersection line is degenerate, and it is not added to the set of intersection lines.
 The intersection between overlapping and co-planar triangles is more complicated; \code{vtkPolyDataIntersection} does not explicitly handle intersections between two coplanar triangles.
@@
 Because the end points of the intersection lines are merged during construction, they need to be split at locations where the mesh is split.
-For example, a mesh representing a cube often has duplicate points at the corners to store three different normals, one for each face incident to the point.
+For example, as shown in Figure \ref{fig:BoxVectors}, a mesh representing a cube often has duplicate points at the corners to store three different normals, one for each face incident to the point.
 These normals enable the appearance of sharp edges on the mesh.
 If the surface intersection on a cube splits an edge that connects two of these corners, then two points should be inserted at the split point rather than one to ensure that two face normals are defined at the split point.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.25]{Figures/BoxNormals}
+\itkcaption{Splitting a mesh with duplicate normals.
+The box object has three duplicate points at each corner each of which has a different normal (gray arrows).
+The surface intersection between the box in the front (gray, solid) and the box in the back (white, outline) is shown in green.
+Points from the intersection lines that lie on the duplicate edges of the box are also duplicated by the mesh splitting procedure in \code{vtkPolyDataIntersection}.
+The normal at a duplicated point, along with other point data, is interpolated from the cell that contains the intersection line that references the point.}
+\label{fig:BoxNormals}
+\end{figure}
 
@@
 
-Splitting proceeds on a cell-by-cell basis. For each cell that needs splitting, the following data is gathered:
+Splitting proceeds on a cell-by-cell basis. For each cell that needs splitting, the following points and lines are gathered:
 
 \begin{enumerate}
-\item The three points that define the cell.
-\item The boundary lines of the cell.
-\item The split intersection lines that lie in the cell (if any).
-\item The points that define the lines in 3.
-\item Points from neighboring cells that lie on a cell edge but which were not already added in 4.
+\item The three points that define the cell (green spheres in Figure \ref{fig:CellSplitSetup}).
+\item The boundary lines of the cell (red lines in Figure \ref{fig:CellSplitSetup}).
+\item The split intersection lines that lie in the cell (if any) (white lines in Figure \ref{fig:CellSplitSetup}).
+\item The points that define the lines in 3 (purple spheres in Figure \ref{fig:CellSplitSetup}).
+\item Points from neighboring cells that lie on a cell edge but which were not already added in 4 (orange sphere in Figure \ref{fig:CellSplitSetup}).
 \end{enumerate}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.25]{Figures/CellSplitSetup}
+\caption{The points and lines required for splitting a cell.}
+\label{fig:CellSplitSetup}
+\end{figure}
 

### 2011-05-10 Document/LaTeX/Article.tex
Added results and discussion about how to use classes to compute surfaces from boolean operations.

@@
 \usepackage[dvips]{graphicx}
+\usepackage{subfigure}
 
@@
 Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seamlessly.
-XXXX Figure
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.25]{Figures/DistanceZeroLevelVsIntersection}
+\itkcaption{A case where the zero level of the distance field from the cylinder (black lines) does not match the intersection between the cone and the cylinder (white lines).
+Because of this mismatch, the class \code{vtkClipPolyData} is not suitable for accurate splitting of one surface mesh by another.}
+\label{fig:DistanceZeroLevelVsIntersection}
+\end{figure}
 
@@
 
+\section{Putting it All Together}
+
+To compute a surface resulting from a boolean operation on surface meshes, the following pipeline of classes can be used to compute the desired surface:
+
+\begin{center}
+\code{vtkPolyDataIntersection}
+
+$\downarrow$
+
+\code{vtkPolyDataDistance}
+
+$\swarrow$ \hspace{3em} $\searrow$
+
+\code{vtkThreshold} \code{vtkThreshold}
+
+$\downarrow$ \hspace{5em}  $\downarrow$
+
+\code{vtkDataSetSurfaceFilter} \code{vtkDataSetSurfaceFilter}
+
+$\searrow$ \hspace{3em} $\swarrow$
+
+\code{vtkAppendPolyData}
+\end{center}
+
+The two instances of \code{vtkThreshold} and two instances of \code{vtkDataSetSurfaceFilter} are required to select different portions from each of the input geometries.
+
+For convenience, we have included a single class that replicates the functionality  class for computing boolean operations on surface meshes, \code{vtkPolyDataBooleanOperationFilter}, that encapsulates the functionality of the pipeline above.
+
 \section{Results}
 
-XXXX Show pictures of various objects with boolean operations applied to them.
+Results from boolean operations on various pairs of geometric objects are shown in Figure \ref{fig:Results}.
+
+\input{ResultsFigure.tex}
 
@@
 
-The source code contributed with this article has been build against VTK 5.8.
+The source code contributed with this article has been built against VTK 5.8.
 

### 2011-05-10 Document/LaTeX/Article.tex
Added docs on vtkImplicitPolyData.

@@
 The method \code{double EvaluateFunction(double x[3])} is overridden to return the signed distance, and \code{void EvaluateGradient(double x[3], double g[3])} is overridden to return the angle-weighted pseudonormal.
+
+The class \code{vtkImplicitPolyData} has several options that can be set. Use
+\begin{verbatim}
+void SetNoValue(double value);
+\end{verbatim}
+to set the value returned by \code{double EvaluateFunction(double x[3])} when an error occurs. Likewise,
+\begin{verbatim}
+void SetNoGradient(double value[3]);
+\end{verbatim}
+to set the gradient returned when an error is encountered. Finally,
+\begin{verbatim}
+void SetTolerance(double tolerance);
+\end{verbatim}
+is used to determine when the absolute value of the signed distance is close enough to zero to be considered zero.
 
@@
 
-\input{ResultsFigure.tex}
+%\input{ResultsFigure.tex}
 

### 2011-05-10 Document/LaTeX/Article.tex
Add docs on methods in vtkPolyDataDistance.

@@
 
+\subsection{vtkImplicitPolyData}
+
 The class \code{vtkImplicitPolyData} has several options that can be set. Use
@@
 
-To support boolean operations, we introduce another class, \code{vtkPolyDataDistance}, that computes the distance from the first input \code{vtkPolyData} to a second.
-Optionally, the distance field from the second input \code{vtkPolyData} to the first can also be computed.
-These distances are computed at points in the mesh and are stored as a point data field named ``Distance''.
+\subsection{vtkPolyDataDistance}
+
+To support distance-based boolean operations, we introduce another class, \code{vtkPolyDataDistance}, that computes the distance from points in the first input \code{vtkPolyData} to a second by evaluating the signed distance field from the second input using the \code{vtkImplicitPolyData} class.
+Optionally, the distance from points in the second input \code{vtkPolyData} to the first can also be computed.
+These distances are stored as a point data field named ``Distance''.
+
+\code{vtkPolyDataDistance} has some options. Use the methods
+\begin{verbatim}
+void SetSignedDistance(int value);
+void SignedDistanceOn();
+void SignedDistanceOff();
+\end{verbatim}
+to turn computation of the signed distance on or off. By default, this option is on. If it is off, the unsigned distance function is computed.
+
+The methods
+\begin{verbatim}
+void SetNegateDistance(int value);
+void NegateDistanceOn();
+void NegateDistanceOff();
+\end{verbatim}
+can be used to enable or disable negation of the signed distance field. If the \code{SignedDistance} option is off, then this option has no effect. \code{NegateDistance} option is off by default.
+
+The methods
+\begin{verbatim}
+void SetComputeSecondDistance(int value);
+void ComputeSecondDistanceOn();
+void ComputeSecondDistanceOff();
+\end{verbatim}
+enables computation of the signed distance for the second input. Finally, a convenience method
+\begin{verbatim}
+vtkPolyData* GetSecondDistanceOutput();
+\end{verbatim}
+can be called to get the second output.
 

### 2011-05-10 Document/LaTeX/Article.tex
More documentation of class methods.

@@
 
-While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, there is a problem.
+While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, doing so would produce inaccurate results.
 \code{vtkClipPolyData} treats the implicit function as piecewise linear within each cell of the \code{vtkPolyData} being clipped.
@@
 Consequently, the clipped portions of each input geometry will not match at the zero level where they should fuse seamlessly.
+Figure \ref{fig:DistanceZeroLevelVsIntersection} shows an example where the zero level of the distance field (black) does not match the surface intersection (white).
 
@@
 
-Instead of directly clipping the input meshes based on the signed distance evaluated at mesh points, it is necessary to split each input mesh at the actual intersection of the two surfaces (hereafter referred to as the \emph{surface intersection}).
+Instead of clipping the input meshes based on the signed distance evaluated at mesh points, it is necessary to split each input mesh at the actual intersection of the two surfaces (hereafter referred to as the \emph{surface intersection}).
 All points on the zero-level of the distance field are then completely defined by the points and edges of the surface intersection, and the surface intersection marks the accurate boundary between inside and outside portions of the mesh surface.
-We have defined a new class \code{vtkPolyDataIntersection} that computes the surface intersection between two surface meshes and can optionally split the input meshes by the surface intersection.
-Its operation is described below.
 
@@
 
+\subsection{vtkPolyDataIntersection}
+
+We have defined a new class \code{vtkPolyDataIntersection} that computes the surface intersection between two surface meshes and can optionally split the input meshes by the surface intersection.
+The surface meshes must consist of triangles.
+Two options are available for this filter:
+\begin{verbatim}
+void SetSplitFirstOutput(int value);
+void SplitFirstOutputOn();
+void SplitFirstOutputOff();
+
+void SetSplitSecondOutput(int value);
+void SplitSecondOutputOn();
+void SplitSecondOutputOff();
+\end{verbatim}
+These options specify whether the output meshes should be split. Each option is on by default.
+
+XXXX Change this in the source code documentation.
+
 \section{Putting it All Together}
 
-To compute a surface resulting from a boolean operation on surface meshes, the following pipeline of classes can be used to compute the desired surface:
+There are two ways to compute the surface mesh resulting from a boolean operation on two surface meshes.
 
+\subsection{Setting up a VTK Pipeline}
+
+To compute a surface resulting from a boolean operation on surface meshes, the pipeline of classes shown in Figure \ref{fig:Pipeline} can be used to compute the desired surface.
+The two instances of \code{vtkThreshold} and two instances of \code{vtkDataSetSurfaceFilter} are required to select different portions from each of the input geometries.
+
+\begin{figure}[h]
 \begin{center}
@@
 \end{center}
+\itkcaption{Possible pipeline to be used to compute results of boolean operations on surface meshes.}
+\label{fig:Pipeline}
+\end{figure}
 
-The two instances of \code{vtkThreshold} and two instances of \code{vtkDataSetSurfaceFilter} are required to select different portions from each of the input geometries.
+\subsection{vtkPolyDataBooleanOperationFilter}
 
 For convenience, we have included a single class that replicates the functionality  class for computing boolean operations on surface meshes, \code{vtkPolyDataBooleanOperationFilter}, that encapsulates the functionality of the pipeline above.
+This class has three options.
+The desired boolean operation is specified with one of
+\begin{verbatim}
+void SetOperationToUnion();
+void SetOperationToIntersection();
+void SetOperationToDifference();
+void SetOperation(int operation);
+\end{verbatim}
+The default operation is union.
 
+When computing the difference boolean operation, part of the resulting surface mesh will have reversed normals and a reversed order of cell points.
+The ReorientDifferenceCells option is used to enable reversal of the normals and reorientation of these cells and can be set using the methods
+\begin{verbatim}
+void SetReorientDifferenceCells(int value);
+void ReorientDifferenceCellsOn();
+void ReorientDifferenceCellsOff();
+\end{verbatim}
+
+Lastly, the Tolerance sets a threshold on the absolute value of a distance field value below which the distance is considered to be zero.
 \section{Results}

### 2011-05-11 Document/LaTeX/Article.tex
Reworked section on intersection line splitting.

@@
 \noindent
-We have written a set of classes than enable boolean operations on the space contained within surface meshes defined in \code{vtkPolyData} objects using only VTK classes.
-In addition to being compatible with the VTK license, our contribution preserves mesh topology and passes point data and cell data through to outputs.
+We have written a set of classes than enable computation of boolean operations on surface meshes using only VTK classes.
+In addition to being compatible with the VTK license, our contribution preserves surface mesh topology (to the extent possible) and passes point data and cell data through to the output mesh where possible.
 
@@
 Points from the intersection lines that lie on the duplicate edges of the box are also duplicated by the mesh splitting procedure in \code{vtkPolyDataIntersection}.
-The normal at a duplicated point, along with other point data, is interpolated from the cell that contains the intersection line that references the point.}
+The normal at a duplicated point, along with other point data, is interpolated from the cell that contains the intersection line that references that point.}
 \label{fig:BoxNormals}
@@
 
-XXXX work on below paragraph again
+More generally, the end result of splitting the surface intersection should be that a copy of each line end point is produced for each set of triangles that share an edge.
+This is done by iterating over the entries of the edge point multimap.
+The first time a particular point ID is encountered in a multimap key-value pair, all other entries with the same key are checked to determine if the cell with the cell ID in the value shares the edge indicated by the first entry's value.
+If the entry is for a shared edge, then the entry is removed from the multimap and not processed any further.
+If not, the entry is left in the multimap for further processing.
 
-More generally, the end result of splitting the surface intersection should be that a copy of each line end point is produced for each set of triangles that share an edge.
-This can be achieved by iterating over the key-value pairs in the multimap, appending a copy of the point corresponding to the point ID in the key to the end of the \code{vtkPoints} object storing the endpoints, and removing all records in the multimap where the point ID is the same as the first point and the cell ID marks a cell that shares the edge identified in the first record.
-The point indices of the lines within the cells that share an edge are renumbered to the ID of the new point in the \code{vtkPoints} object.
-Finally, the point ID and the cell ID from one of the cells that shares the edge is stored in a point-to-cell map for future use in interpolating point data at the new point.
-It does not matter which cell ID is associated with a point on the edge of a cell because the interpolation results will be the same.
+If any multimap entries are left that have the current point ID as key, then that point lies on a split in the mesh and requires duplication.
+In this case, a copy of the point is inserted into the list of points for the intersection line data structure, and the first such entry in the multimap is read.
+All other entries in the multimap with the point ID as key are again examined to determine which share an edge with the first entry.
+Entries that do not share an edge are left in the multimap.
+For entries that share an edge, the point ID in the line referenced by the line ID stored in the entry is replaced with the ID of the newly duplicated point.
+The entry is then removed from the multimap.
+The process described in this paragraph is repeated if any entries are left that have the current point ID as key.
+Otherwise, the first multimap entry is removed.
 
-The points in the input mesh and their associated point data are copied to the points in the output mesh.
+During the above process, another auxiliary data structure is built.
+Every time a point is duplicated, a key-value pair is added to a map where the key is the point ID and the value is the cell ID from which the point data at the point should be interpolated.
+It does not matter which cell ID from the set of cells that share an edge is associated with a point on that edge because the interpolation results will be the same.
+
+After the intersection lines are split, the points in the input mesh and their associated point data are copied to the points in the output mesh.
 The points from the intersection lines are then appended to the point set in the output mesh, and the point data is interpolated at those points.
@@
 
-\subsection{Setting up a VTK Pipeline}
+\subsection{Setting up a VTK Pipeline for Boolean Operations on Surfaces}
 
@@
 
-\code{vtkThreshold} \code{vtkThreshold}
+\code{vtkThreshold} \hspace{1em} \code{vtkThreshold}
 
-$\downarrow$ \hspace{5em}  $\downarrow$
+$\downarrow$ \hspace{6em}  $\downarrow$
 
-\code{vtkDataSetSurfaceFilter} \code{vtkDataSetSurfaceFilter}
+\code{vtkDataSetSurfaceFilter} \hspace{1em} \code{vtkDataSetSurfaceFilter}
 
@@
 When computing the difference boolean operation, part of the resulting surface mesh will have reversed normals and a reversed order of cell points.
-The ReorientDifferenceCells option is used to enable reversal of the normals and reorientation of these cells and can be set using the methods
+The \code{ReorientDifferenceCells} option is used to enable reversal of the normals and reorientation of these cells and can be set using the methods
 \begin{verbatim}
@@
 
-Lastly, the Tolerance sets a threshold on the absolute value of a distance field value below which the distance is considered to be zero.
+Lastly, the \code{Tolerance} sets a threshold on the absolute value of a distance field value below which the distance is considered to be zero.
 \section{Results}

### 2011-05-11 Document/LaTeX/Article.tex
Converging to a final draft.

@@
 This article describes an implementation of those methods to realize boolean operations on surfaces in VTK.
+
+\section{Previous Work}
+
+Lloyd previously contributed a VTK class for computing boolean operations on surface meshes.
+His contribution wrapped the GNU Triangulated Surface Library (GTS) in a VTK class that performed conversion between the VTK mesh data structure and the GTS mesh data structure and vice versa.
+This work filled a need often requested by users on the VTK mailing lists.
+
+Unfortunately, his work is unable to be incorporated into the main VTK repository because of the  incompatible GTS license.
+Furthermore, any projects using Lloyd's contribution are subject to the licensing requirements of GTS.
+Our contribution is not subject to an incompatible license as it makes use of no external libraries or code.
+
+Furthermore, our contribution copies point and cell data from the input meshes to the output meshes appropriately, interpolating point data where necessary.
+When portions of two input meshes are combined into one mesh, only the common point and cell data in both meshes is copied to the output.
 
@@
 
-XXXX Change this in the source code documentation.
-
 \section{Putting it All Together}
@@
 
-$\downarrow$
-
-\code{vtkPolyDataDistance}
-
 $\swarrow$ \hspace{3em} $\searrow$
 
-\code{vtkThreshold} \hspace{1em} \code{vtkThreshold}
-
-$\downarrow$ \hspace{6em}  $\downarrow$
-
-\code{vtkDataSetSurfaceFilter} \hspace{1em} \code{vtkDataSetSurfaceFilter}
+\code{vtkImplicitPolyData} $\rightarrow$ \code{vtkClipPolyData} \hspace{1em} \code{vtkClipPolyData} $\leftarrow$ \code {vtkImplicityPolyData}
 
@@
 
-%\input{ResultsFigure.tex}
+\input{ResultsFigure.tex}
 

### 2011-05-11 Source/vtkImplicitPolyData.h
Added source files.

@@ -1,0 +1,101 @@
+/*=========================================================================
+
+  Program:   Visualization Toolkit
+  Module:    vtkImplicitPolyData.h
+
+  Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
+  All rights reserved.
+  See Copyright.txt or http://www.kitware.com/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notice for more information.
+
+=========================================================================*/
+// .NAME vtkImplicitPolyData
+// .SECTION Description
+//
+// Implicit function that computes the distance from a point x to the
+// nearest point p on an input vtkPolyData. The sign of the function
+// is set to the sign of the dot product between the angle-weighted
+// pseudo-normal at the nearest surface point and the vector x -
+// p. The gradient of the function is the angle-weighted pseudo-normal
+// at the neared point.
+//
+// Written by Chris Weigle and Cory Quammen, The University of North
+// Carolina at Chapel Hill.
+
+#ifndef __vtkImplicitPolyData_h
+#define __vtkImplicitPolyData_h
+
+#include "vtkImplicitFunction.h"
+
+
+class vtkCellLocator;
+class vtkPolyData;
+class vtkTriangleFilter;
+
+class vtkImplicitPolyData : public vtkImplicitFunction
+{
+public:
+  static vtkImplicitPolyData *New();
+  vtkTypeRevisionMacro(vtkImplicitPolyData,vtkImplicitFunction);
+  void PrintSelf(ostream& os, vtkIndent indent);
+
+  // Description:
+  // Return the MTime also considering the Input dependency.
+  unsigned long GetMTime();
+
+  // Description:
+  // Evaluate plane equation of nearest triangle to point x[3].
+  double EvaluateFunction(double x[3]);
+
+  // Description:
+  // Evaluate function gradient of nearest triangle to point x[3].
+  void EvaluateGradient(double x[3], double g[3]);
+
+  // Description:
+  // Set the input vtkPolyData used for the implicit function
+  // evaluation.  Passes input through an internal instance of
+  // vtkTriangleFilter to remove vertices and lines, leaving only
[…]

### 2011-05-11 Source/vtkPolyDataBooleanOperationFilter.h

@@ -1,0 +1,123 @@
+/*=========================================================================
+
+  Program:   Visualization Toolkit
+  Module:    vtkPolyDataBooleanOperationFilter.h
+
+  Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
+  All rights reserved.
+  See Copyright.txt or http://www.kitware.com/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notice for more information.
+
+=========================================================================*/
+// .NAME vtkPolyDataBooleanOperationFilter
+// .SECTION Description
+//
+// Computes the boundary of the union, intersection, or difference
+// volume computed from the volumes defined by two input surfaces. The
+// two surfaces do not need to be manifold. The resulting surface is
+// available in the first output of the filter. The second output
+// contains a set of polylines that represent the intersection between
+// the two input surfaces.
+
+#ifndef __vtkPolyDataBooleanOperationFilter_h
+#define __vtkPolyDataBooleanOperationFilter_h
+
+#include <vtkPolyDataAlgorithm.h>
+
+#include <vtkDataSetAttributes.h> // Needed for CopyCells() method
+
+class vtkIdList;
+class vtkPolyDataDistance;
+class vtkPolyDataIntersection;
+
+
+class vtkPolyDataBooleanOperationFilter : public vtkPolyDataAlgorithm
+{
+public:
+  // Description:
+  // Construct object that computes the boolean surface.
+  static vtkPolyDataBooleanOperationFilter *New();
+
+  vtkTypeRevisionMacro(vtkPolyDataBooleanOperationFilter,
+                       vtkPolyDataAlgorithm);
+
+  void PrintSelf(ostream& os, vtkIndent indent);
+
+  enum OperationTypes
+  {
+    UNION=0,
+    INTERSECTION,
+    DIFFERENCE
+  };
+
+  vtkSetClampMacro( Operation, int, UNION, DIFFERENCE );
+  vtkGetMacro( Operation, int );
+  void SetOperationToUnion()
+  { this->SetOperation( UNION ); }
+  void SetOperationToIntersection()
+  { this->SetOperation( INTERSECTION ); }
+  void SetOperationToDifference()
+  { this->SetOperation( DIFFERENCE ); }
+
+  // Description:
[…]

### 2011-05-11 Source/vtkPolyDataDistance.h

@@ -1,0 +1,74 @@
+/*=========================================================================
+
+  Written by: Chris Weigle, University of North Carolina at Chapel Hill
+  Email Contact: weigle@cs.unc.edu
+
+  IN NO EVENT SHALL THE UNIVERSITY OF NORTH CAROLINA AT CHAPEL HILL BE
+  LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL, INCIDENTAL, OR
+  CONSEQUENTIAL DAMAGES, INCLUDING LOST PROFITS, ARISING OUT OF THE USE
+  OF THIS SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE UNIVERSITY OF NORTH
+  CAROLINA HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGES.
+
+  THE UNIVERSITY OF NORTH CAROLINA SPECIFICALLY DISCLAIM ANY WARRANTIES,
+  INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY
+  AND FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
+  ON AN "AS IS" BASIS, AND THE UNIVERSITY OF NORTH CAROLINA HAS NO
+  OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
+  MODIFICATIONS.
+
+=========================================================================*/
+#ifndef __vtkPolyDataDistance_h
+#define __vtkPolyDataDistance_h
+
+#include <vtkPolyDataAlgorithm.h>
+
+class vtkPolyDataDistance : public vtkPolyDataAlgorithm {
+public:
+  static vtkPolyDataDistance *New();
+  vtkTypeRevisionMacro(vtkPolyDataDistance, vtkPolyDataAlgorithm);
+  void PrintSelf(ostream& os, vtkIndent indent);
+
+  // Description:
+  // Enable/disable computation of the signed distance between
+  // the first poly data and the second poly data.
+  vtkSetMacro(SignedDistance,int);
+  vtkGetMacro(SignedDistance,int);
+  vtkBooleanMacro(SignedDistance,int);
+
+  // Description:
+  // Enable/disable negation of the distance values.
+  vtkSetMacro(NegateDistance,int);
[…]

### 2011-05-11 Source/vtkPolyDataIntersection.h

@@ -1,0 +1,83 @@
+/*=========================================================================
+
+  Program:   Visualization Toolkit
+  Module:    vtkPolyDataIntersection.h
+
+  Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
+  All rights reserved.
+  See Copyright.txt or http://www.kitware.com/Copyright.htm for details.
+
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notice for more information.
+
+=========================================================================*/
+// .NAME vtkPolyDataIntersection
+// .SECTION Description
+//
+// vtkPolyDataIntersection computes the intersection between two
+// vtkPolyData objects. The first output is a set of lines that marks
+// the intersection of the input vtkPolyData objects. The second and
+// third outputs are the first and second input vtkPolyData,
+// respectively. Optionally, the two output vtkPolyData can be split
+// along the intersection lines.
+//
+// Written by Chris Weigle and Cory Quammen, The University of North
+// Carolina at Chapel Hill.
+
+#ifndef __vtkPolyDataIntersection_h
+#define __vtkPolyDataIntersection_h
+
+#include "vtkPolyDataAlgorithm.h"
+
+
+class vtkPolyDataIntersection : public vtkPolyDataAlgorithm
+{
+public:
+  static vtkPolyDataIntersection *New();
+  vtkTypeMacro(vtkPolyDataIntersection, vtkPolyDataAlgorithm);
+  virtual void PrintSelf(ostream &os, vtkIndent indent);
+
+  // Description:
+  // If on, the second output will be the first input mesh split by the
+  // intersection with the second input mesh. Defaults to on.
+  vtkGetMacro(SplitFirstOutput, int);
+  vtkSetMacro(SplitFirstOutput, int);
+  vtkBooleanMacro(SplitFirstOutput, int);
+
+  // Description:
+  // If on, the third output will be the second input mesh split by the
+  // intersection with the first input mesh. Defaults to on.
+  vtkGetMacro(SplitSecondOutput, int);
+  vtkSetMacro(SplitSecondOutput, int);
+  vtkBooleanMacro(SplitSecondOutput, int);
+
+  // Description:
+  // Given two triangles defined by points (p1, q1, r1) and (p2, q2,
+  // r2), returns whether the two triangles intersect. If they do,
[…]

### 2011-05-11 Document/LaTeX/Article.tex
Changed documentation on VTK pipeline to old version after investigations into using vtkClipPolyData didn't pan out (it merges points).

@@
 
-There are two ways to compute the surface mesh resulting from a boolean operation on two surface meshes.
+There are two ways to compute the surface mesh resulting from a boolean operation on two surface meshes: setting up a VTK pipeline or using a new class \code{vtkPolyDataBooleanOperationFilter}.
 
@@
 To compute a surface resulting from a boolean operation on surface meshes, the pipeline of classes shown in Figure \ref{fig:Pipeline} can be used to compute the desired surface.
-The two instances of \code{vtkThreshold} and two instances of \code{vtkDataSetSurfaceFilter} are required to select different portions from each of the input geometries.
+The two instances of \code{vtkThreshold} are required to select different portions from each of the input geometries.
 
@@
 
+$\downarrow$
+
+\code{vtkPolyDataDistance}
+
 $\swarrow$ \hspace{3em} $\searrow$
 
-\code{vtkImplicitPolyData} $\rightarrow$ \code{vtkClipPolyData} \hspace{1em} \code{vtkClipPolyData} $\leftarrow$ \code {vtkImplicityPolyData}
+\code{vtkThreshold} \hspace{1em} \code{vtkThreshold}
 
@@
 
-\code{vtkAppendPolyData}
+\code{vtkAppendFilter}
+
+$\downarrow$
+
+\code{vtkDataSetSurfaceFilter}
 \end{center}
@@
 The \code{vtkPolyDataIntersection} class operates only on triangulated surface meshes, and it does not properly handle intersections between coplanar triangles.
-When such an intersection event has been detected, the intersection information is discarded.
-If this occurs, then the output of a boolean operation may not be exactly correct at the location of the co-planar triangles.
+When two triangles are determined to be coplanar, the intersection information is discarded.
+If this occurs, then the output of a boolean operation may not be exactly correct at the location of the coplanar triangles.
 
@@
 
-The source code contributed with this article has been built against VTK 5.8.
+The source code contributed with this article has been built against VTK git commit 9f132a45a7bece15d77ad74723c6378f2f29d869.
 

### 2011-05-12 Document/LaTeX/Article.tex
More edits.

@@
 % the relevant document. 
-\title{Boolean Operations for Surfaces in VTK Without External Libraries}
+\title{Boolean Operations on Surfaces in VTK Without External Libraries}
 
@@
 We have written a set of classes than enable computation of boolean operations on surface meshes using only VTK classes.
-In addition to being compatible with the VTK license, our contribution preserves surface mesh topology (to the extent possible) and passes point data and cell data through to the output mesh where possible.
+In addition to being compatible with the VTK license, our contribution preserves surface mesh topology to the extent possible in boolean operations and passes point data and cell data through to the output mesh where possible.
 
@@
 In geometric modeling, a closed, orientable 2-manifold surface mesh in 3D $M_A$ may be considered the boundary of an infinite set of 3D points $A$.
-Given a second surface mesh $M_B$ with the same properties, a boolean operation may be applied to obtain a third surface mesh $M_C$.
+Given a second surface mesh $M_B$ with the same properties, a boolean operation may be applied to obtain a third surface mesh $M_C$ that bounds the infinite set of points $C$ resulting from the boolean operation on $A$ and $B$.
+For brevity, we use the phrase ``boolean operation on surfaces'' to refer to the process of determining $M_C$ from $M_A$, $M_B$, and a boolean operation.
 These simple operations can be used to define complex geometries.
 
-Rather than converting two mesh representations to a set of points, performing the boolean operations on those points, and reconstructing a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
+Rather than somehow converting two mesh representations to a set of points, performing the boolean operations on those points, and reconstructing a third mesh from the resulting points, direct operations on the bounding geometry based on signed distance fields and mesh-to-mesh clipping can be used.
 This article describes an implementation of those methods to realize boolean operations on surfaces in VTK.
@@
 
-Lloyd previously contributed a VTK class for computing boolean operations on surface meshes.
+Lloyd previously contributed a VTK class for computing boolean operations on surface meshes \cite{Lloyd2010}.
 His contribution wrapped the GNU Triangulated Surface Library (GTS) in a VTK class that performed conversion between the VTK mesh data structure and the GTS mesh data structure and vice versa.
@@
 
-Furthermore, our contribution copies point and cell data from the input meshes to the output meshes appropriately, interpolating point data where necessary.
-When portions of two input meshes are combined into one mesh, only the common point and cell data in both meshes is copied to the output.
+Furthermore, our contribution extend's Lloyd's work by copying point and cell data from the input meshes to the output meshes appropriately, interpolating point data where necessary.
+When portions of two input meshes are combined into one mesh, the common point and cell data in both meshes is copied to the output.
 
@@
 
-The class \code{vtkImplicitPolyData} has several options that can be set. Use
+The class \code{vtkImplicitPolyData} operates on a \code{vtkPolyData} input. Use the method
+\begin{verbatim}
+void SetInput(vtkPolyData *input);
+\end{verbatim}
+to specify the \code{vtkPolyData} on which it should operate. The class also has several options that can be set. Use
 \begin{verbatim}
@@
 \end{verbatim}
-is used to determine when the absolute value of the signed distance is close enough to zero to be considered zero.
+use to determine when the absolute value of the signed distance is close enough to zero to be considered zero.
 
@@
 
-The methods
+Use the methods
 \begin{verbatim}
@@
 \end{verbatim}
-can be used to enable or disable negation of the signed distance field. If the \code{SignedDistance} option is off, then this option has no effect. \code{NegateDistance} option is off by default.
+to enable or disable negation of the signed distance field. If the \code{SignedDistance} option is off, then this option has no effect. \code{NegateDistance} option is off by default.
 
@@
 Boolean operations can be computed with only the signed distance field for each mesh.
-The sign of the distance field at a point in the mesh corresponds to whether that point is inside (negative), outside (positive), or on (zero) the other mesh; the surfaces bounding the volumes produced by the boolean operations described in Section \ref{sec:Introduction} can thus be defined as:
+The sign of the distance field at a point in the mesh corresponds to whether that point is inside (negative), outside (positive), or on (zero) the other mesh; the surfaces bounding the volumes produced by the boolean operations described in Section \ref{sec:Introduction} can therefore be defined as:
 
@@
 While it seems possible to use the class \code{vtkClipPolyData} together with \code{vtkImplicitPolyData} to extract the necessary portions of each input surface to produce the boolean surfaces, doing so would produce inaccurate results.
-\code{vtkClipPolyData} treats the implicit function as piecewise linear within each cell of the \code{vtkPolyData} being clipped.
-In general, implicit functions are not guaranteed to be piecewise linear in each cell, and the same is true for the signed distance field.
+\code{vtkClipPolyData} treats the implicit function as a single piecewise linear function within each cell of the \code{vtkPolyData} being clipped.
+In general, an implicit function is not guaranteed to be a single piecewise linear function in each cell, and the same is true for the signed distance field.
 The result is that clipped surfaces produced by \code{vtkClipPolyData} will not be clipped at the true zero level of the distance field.
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \centering
 \includegraphics[scale=0.25]{Figures/DistanceZeroLevelVsIntersection}
-\itkcaption{A case where the zero level of the distance field from the cylinder (black lines) does not match the intersection between the cone and the cylinder (white lines).
-Because of this mismatch, the class \code{vtkClipPolyData} is not suitable for accurate splitting of one surface mesh by another.}
+\itkcaption{A case where the zero level of the distance field from the cylinder (black lines) computed on the cone does not match the intersection between the cylinder and the cone (white lines).
+Because of this mismatch, the class \code{vtkClipPolyData} is not suitable for accurate splitting of one surface mesh by another without further mesh processing.}
 \label{fig:DistanceZeroLevelVsIntersection}
@@
 
-Identifying the surface intersection between two surface meshes is the most complex part of the boolean operations algorithm.
+Identifying the surface intersection between two surface meshes is the most complex part of the boolean operation algorithm.
 It involves identifying the intersection between each triangle in the first mesh with all triangles in the second mesh that intersect it.
@@
 If the endpoints are the same, then the intersection line is degenerate, and it is not added to the set of intersection lines.
-The intersection between overlapping and co-planar triangles is more complicated; \code{vtkPolyDataIntersection} does not explicitly handle intersections between two coplanar triangles.
+The intersection between overlapping and co-planar triangles is more complicated; our work does not explicitly handle intersections between two coplanar triangles.
 
@@
 The method \code{vtkOBBTree::IntersectWithOBBTree()} is then called on one OBB tree with the second OBB tree and a callback function passed as parameters.
-The callback function performs more accurate triangle-triangle intersection tests between the triangles in overlapping nodes from the two OBB trees and stores the results in several data structures used later in the algorithm.
+The callback function performs a more accurate triangle-triangle intersection test between the triangles in overlapping nodes from the two OBB trees and stores the line defining the intersection and other information in several data structures used later in the algorithm.
 
@@
 Merging the end points also simplifies later operations.
-In addition, each end point is tested to see if it lies on each edge of the two input triangles.
+In addition, each end point is tested to see if it lies on an edge of the two input triangles.
 If so, a key-value pair is stored in a \code{std::multimap} where the key is the ID of the endpoint and the value is a 3-tuple consisting of the index of the triangle, the index of the triangle edge, and the index of the intersecting line.
@@
 The surface intersection between the box in the front (gray, solid) and the box in the back (white, outline) is shown in green.
-Points from the intersection lines that lie on the duplicate edges of the box are also duplicated by the mesh splitting procedure in \code{vtkPolyDataIntersection}.
+Points from the intersection lines that lie on the duplicate edges of the box are also duplicated by the mesh splitting procedure.
 The normal at a duplicated point, along with other point data, is interpolated from the cell that contains the intersection line that references that point.}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 
-We have defined a new class \code{vtkPolyDataIntersection} that computes the surface intersection between two surface meshes and can optionally split the input meshes by the surface intersection.
+We have defined a new class \code{vtkPolyDataIntersection} that encapsulates the algorithm described above to compute the surface intersection between two surface meshes and optionally split the input meshes by the surface intersection.
 The surface meshes must consist of triangles.
@@
 
-There are two ways to compute the surface mesh resulting from a boolean operation on two surface meshes: setting up a VTK pipeline or using a new class \code{vtkPolyDataBooleanOperationFilter}.
+There are two ways to use the classes we have developed to compute the surface mesh resulting from a boolean operation on two surface meshes: setting up a VTK pipeline or using a new class \code{vtkPolyDataBooleanOperationFilter}.
 
@@
 To compute a surface resulting from a boolean operation on surface meshes, the pipeline of classes shown in Figure \ref{fig:Pipeline} can be used to compute the desired surface.
-The two instances of \code{vtkThreshold} are required to select different portions from each of the input geometries.
+The two arrows between \code{vtkPolyDataIntersection} and \code{vtkPolyDataDistance} indicate that the two output surfaces from \code{vtkPolyDataIntersection} are set as the input to \code{vtkPolyDataDistance}.
+The two instances of \code{vtkThreshold} operate on different outputs of \code{vtkPolyDataDistance}.
 
-\begin{figure}[h]
+\begin{figure}[t]
 \begin{center}
+
+\code{vtkPolyData} \hspace{1em} \code{vtkPolyData}
+
+$\downarrow$ \hspace{3.2em} $\downarrow$
+
 \code{vtkPolyDataIntersection}
 
-$\downarrow$
+$\downarrow$ \hspace{3.2em} $\downarrow$
 
@@
 
-$\searrow$ \hspace{3em} $\swarrow$
+$\downarrow$ \hspace{4em} $\downarrow$
 
-\code{vtkAppendFilter}
+\code{vtkDataSetSurface} \hspace{1em} \code{vtkDataSetSurface}
+
+$\downarrow$ \hspace{4em} $\downarrow$
+
+\hspace{5.5em} $\downarrow$ \hspace{2.6em} \code{vtkReverseSense}
+
+$\downarrow$ \hspace{4em} $\downarrow$
+
+\code{vtkAppendPolyDataFilter}
 
@@
 \end{center}
-\itkcaption{Possible pipeline to be used to compute results of boolean operations on surface meshes.}
+\itkcaption{Possible pipeline for computing boolean operations on surface meshes.
+The instance of \code{vtkReverseSense} is optional and can be used when the boolean difference is desired to reorient the intersection portion of the second surface so that its normals face toward the outside of the resulting surface.}
 \label{fig:Pipeline}

### 2011-05-12 Source/vtkImplicitPolyData.h
Another round of edits.

@@ -18,9 +18,15 @@
 // Implicit function that computes the distance from a point x to the
 // nearest point p on an input vtkPolyData. The sign of the function
 // is set to the sign of the dot product between the angle-weighted
-// pseudo-normal at the nearest surface point and the vector x -
-// p. The gradient of the function is the angle-weighted pseudo-normal
-// at the neared point.
+// pseudonormal at the nearest surface point and the vector x - p.
+// Points interior to the geometry have a negative distance, points on
+// the exterior have a positive distance, and points on the input
+// vtkPolyData have a distance of zero. The gradient of the function
+// is the angle-weighted pseudonormal at the nearest point.
+//
+// Baerentzen, J. A. and Aanaes, H. (2005). Signed distance
+// computation using the angle weighted pseudonormal. IEEE
+// Transactions on Visualization and Computer Graphics, 11:243-253.
 //
 // Written by Chris Weigle and Cory Quammen, The University of North
 // Carolina at Chapel Hill.

### 2011-05-12 Source/vtkPolyDataBooleanOperationFilter.h

@@ -17,17 +17,21 @@
 //
 // Computes the boundary of the union, intersection, or difference
 // volume computed from the volumes defined by two input surfaces. The
-// two surfaces do not need to be manifold. The resulting surface is
+// two surfaces do not need to be manifold, but if they are not,
+// unexpected results may be obtained. The resulting surface is
 // available in the first output of the filter. The second output
 // contains a set of polylines that represent the intersection between
 // the two input surfaces.
+//
+// Written by Chris Weigle and Cory Quammen, The University of North
+// Carolina at Chapel Hill.
 
 #ifndef __vtkPolyDataBooleanOperationFilter_h
 #define __vtkPolyDataBooleanOperationFilter_h
 
-#include <vtkPolyDataAlgorithm.h>
+#include "vtkPolyDataAlgorithm.h"
 
-#include <vtkDataSetAttributes.h> // Needed for CopyCells() method
+#include "vtkDataSetAttributes.h" // Needed for CopyCells() method
 
 class vtkIdList;
 class vtkPolyDataDistance;
@@ -53,6 +57,8 @@ public:
     DIFFERENCE
   };
 
+  // Description:
+  // Set the boolean operation to perform. Defaults to union.
   vtkSetClampMacro( Operation, int, UNION, DIFFERENCE );
   vtkGetMacro( Operation, int );
   void SetOperationToUnion()
@@ -64,14 +70,14 @@ public:
 
   // Description:
   // Turn on/off cell reorientation of the intersection portion of the
-  // surface when the operation is set to DIFFERENCE.
+  // surface when the operation is set to DIFFERENCE. Defaults to on.
   vtkSetMacro( ReorientDifferenceCells, int );
   vtkGetMacro( ReorientDifferenceCells, int );
   vtkBooleanMacro( ReorientDifferenceCells, int );
 
   // Description:
   // Set/get the tolerance used to determine when a point's absolute
-  // distance is considered to be zero.
+  // distance is considered to be zero. Defaults to 1e-6.
   vtkSetMacro(Tolerance, double);
   vtkGetMacro(Tolerance, double);
 

### 2011-05-12 Source/vtkPolyDataDistance.h

@@ -1,26 +1,38 @@
 /*=========================================================================
 
-  Written by: Chris Weigle, University of North Carolina at Chapel Hill
-  Email Contact: weigle@cs.unc.edu
+  Program:   Visualization Toolkit
+  Module:    vtkPolyDataDistance.h
 
-  IN NO EVENT SHALL THE UNIVERSITY OF NORTH CAROLINA AT CHAPEL HILL BE
-  LIABLE TO ANY PARTY FOR DIRECT, INDIRECT, SPECIAL, INCIDENTAL, OR
-  CONSEQUENTIAL DAMAGES, INCLUDING LOST PROFITS, ARISING OUT OF THE USE
-  OF THIS SOFTWARE AND ITS DOCUMENTATION, EVEN IF THE UNIVERSITY OF NORTH
-  CAROLINA HAVE BEEN ADVISED OF THE POSSIBILITY OF SUCH DAMAGES.
+  Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
+  All rights reserved.
+  See Copyright.txt or http://www.kitware.com/Copyright.htm for details.
 
-  THE UNIVERSITY OF NORTH CAROLINA SPECIFICALLY DISCLAIM ANY WARRANTIES,
-  INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY
-  AND FITNESS FOR A PARTICULAR PURPOSE. THE SOFTWARE PROVIDED HEREUNDER IS
-  ON AN "AS IS" BASIS, AND THE UNIVERSITY OF NORTH CAROLINA HAS NO
-  OBLIGATION TO PROVIDE MAINTENANCE, SUPPORT, UPDATES, ENHANCEMENTS, OR
-  MODIFICATIONS.
+     This software is distributed WITHOUT ANY WARRANTY; without even
+     the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR
+     PURPOSE.  See the above copyright notice for more information.
 
 =========================================================================*/
+// .NAME vtkPolyDataDistance
+// .SECTION Description
+//
+// Computes the signed distance from one vtkPolyData to another. The
+// signed distance to the second input is computed at every point in
+// the first input using vtkImplicitPolyData. Optionally, the signed
+// distance to the first input at every point in the second input can
+// be computed. This may be enabled by calling
+// ComputeSecondDistanceOn().
+//
[…]

### 2011-05-16 Source/vtkPolyDataDistance.h
Replaced deprecated vtkTypeRevisionMacro with vtkTypeMacro.

@@ -37,7 +37,7 @@
 class vtkPolyDataDistance : public vtkPolyDataAlgorithm {
 public:
   static vtkPolyDataDistance *New();
-  vtkTypeRevisionMacro(vtkPolyDataDistance, vtkPolyDataAlgorithm);
+  vtkTypeMacro(vtkPolyDataDistance, vtkPolyDataAlgorithm);
   void PrintSelf(ostream& os, vtkIndent indent);
 
   // Description:

### 2011-05-16 Source/vtkImplicitPolyData.h
Replaced deprecated vtkTypeRevisionMacro with vtkTypeMacro.

@@ -45,7 +45,7 @@ class vtkImplicitPolyData : public vtkImplicitFunction
 {
 public:
   static vtkImplicitPolyData *New();
-  vtkTypeRevisionMacro(vtkImplicitPolyData,vtkImplicitFunction);
+  vtkTypeMacro(vtkImplicitPolyData,vtkImplicitFunction);
   void PrintSelf(ostream& os, vtkIndent indent);
 
   // Description:

### 2011-05-16 Source/vtkPolyDataBooleanOperationFilter.h
Replaced raw pointers with smart pointers for VTK classes with local scope.
Replaced deprecated vtkTypeRevisionMacro with vtkTypeMacro.

@@ -45,8 +45,8 @@ public:
   // Construct object that computes the boolean surface.
   static vtkPolyDataBooleanOperationFilter *New();
 
-  vtkTypeRevisionMacro(vtkPolyDataBooleanOperationFilter,
-                       vtkPolyDataAlgorithm);
+  vtkTypeMacro(vtkPolyDataBooleanOperationFilter,
+               vtkPolyDataAlgorithm);
 
   void PrintSelf(ostream& os, vtkIndent indent);
 

### 2011-05-18 Source/vtkBooleanOperationPolyDataFilter.h
Renamed vtkPolyDataBooleanOperationFilter to vtkBooleanOperationPolyDataFilter.

@@ -1,7 +1,7 @@
 /*=========================================================================
 
   Program:   Visualization Toolkit
-  Module:    vtkPolyDataBooleanOperationFilter.h
+  Module:    vtkBooleanOperationPolyDataFilter.h
 
   Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
   All rights reserved.
@@ -12,7 +12,7 @@
      PURPOSE.  See the above copyright notice for more information.
 
 =========================================================================*/
-// .NAME vtkPolyDataBooleanOperationFilter
+// .NAME vtkBooleanOperationPolyDataFilter
 // .SECTION Description
 //
 // Computes the boundary of the union, intersection, or difference
@@ -26,8 +26,8 @@
 // Written by Chris Weigle and Cory Quammen, The University of North
 // Carolina at Chapel Hill.
 
-#ifndef __vtkPolyDataBooleanOperationFilter_h
-#define __vtkPolyDataBooleanOperationFilter_h
+#ifndef __vtkBooleanOperationPolyDataFilter_h
+#define __vtkBooleanOperationPolyDataFilter_h
 
 #include "vtkPolyDataAlgorithm.h"
 
@@ -38,14 +38,14 @@ class vtkPolyDataDistance;
 class vtkPolyDataIntersection;
 
 
-class vtkPolyDataBooleanOperationFilter : public vtkPolyDataAlgorithm
+class vtkBooleanOperationPolyDataFilter : public vtkPolyDataAlgorithm
 {
 public:
   // Description:
   // Construct object that computes the boolean surface.
-  static vtkPolyDataBooleanOperationFilter *New();
+  static vtkBooleanOperationPolyDataFilter *New();
 
-  vtkTypeMacro(vtkPolyDataBooleanOperationFilter,
+  vtkTypeMacro(vtkBooleanOperationPolyDataFilter,
                vtkPolyDataAlgorithm);
 
   void PrintSelf(ostream& os, vtkIndent indent);
@@ -82,8 +82,8 @@ public:
   vtkGetMacro(Tolerance, double);
 
 protected:
-  vtkPolyDataBooleanOperationFilter();
-  ~vtkPolyDataBooleanOperationFilter();
+  vtkBooleanOperationPolyDataFilter();
+  ~vtkBooleanOperationPolyDataFilter();
 
   // Description:
   // Tolerance used to determine when a point's absolute
@@ -108,8 +108,8 @@ protected:
   int ReorientDifferenceCells;
 
 private:
[…]

### 2011-05-18 Document/LaTeX/Article.tex
Changed the handle number to the correct one.
Updated the document to reflect the latest changes.

@@
 %
-\newcommand{\IJhandlerIDnumber}{1338}
+\newcommand{\IJhandlerIDnumber}{3262}
 
@@
 % The author and/or editor can define 'significant' however they like.
-\release{1.00}
+\release{2.00}
 
@@
 These distances are stored as a point data field named ``Distance''.
+In addition, the filter computes a cell data field named ``Distance'' that contains the distance from the center of each cell to the nearest point on the second input
 
@@
 
-There are two ways to use the classes we have developed to compute the surface mesh resulting from a boolean operation on two surface meshes: setting up a VTK pipeline or using a new class \code{vtkPolyDataBooleanOperationFilter}.
+There are two ways to use the classes we have developed to compute the surface mesh resulting from a boolean operation on two surface meshes: setting up a VTK pipeline or using a new class \code{vtkBooleanOperationPolyDataFilter}.
 
@@
 The two arrows between \code{vtkPolyDataIntersection} and \code{vtkPolyDataDistance} indicate that the two output surfaces from \code{vtkPolyDataIntersection} are set as the input to \code{vtkPolyDataDistance}.
-The two instances of \code{vtkThreshold} operate on different outputs of \code{vtkPolyDataDistance}.
+The two instances of \code{vtkThreshold} operate on the two outputs of \code{vtkPolyDataDistance}.
+The example \verb+PipelineExample.cxx+ included with the source code in this article shows how to set up a pipeline to compute boolean operations.
 
@@
 
-\subsection{vtkPolyDataBooleanOperationFilter}
+\subsection{vtkBooleanOperationPolyDataFilter}
 
-For convenience, we have included a single class that replicates the functionality  class for computing boolean operations on surface meshes, \code{vtkPolyDataBooleanOperationFilter}, that encapsulates the functionality of the pipeline above.
+For convenience, we have included a single class that replicates the functionality  class for computing boolean operations on surface meshes, \code{vtkBooleanOperationPolyDataFilter}, that encapsulates the functionality of the pipeline above.
 This class has three options.
@@
 Lastly, the \code{Tolerance} sets a threshold on the absolute value of a distance field value below which the distance is considered to be zero.
+
+An example in the file \verb+BooleanOperationPolyDataFilterExample.cxx+ is included in the source code for this article.
+
 \section{Results}

### 2011-05-20 Source/vtkBooleanOperationPolyDataFilter.h
Renamed vtkPolyDataDistance to vtkDistancePolyDataFilter.

@@ -34,7 +34,7 @@
 #include "vtkDataSetAttributes.h" // Needed for CopyCells() method
 
 class vtkIdList;
-class vtkPolyDataDistance;
+class vtkDistancePolyDataFilter;
 class vtkPolyDataIntersection;
 
 
@@ -122,7 +122,7 @@ private:
 
   vtkPolyDataIntersection *PolyDataIntersection;
 
-  vtkPolyDataDistance *PolyDataDistance;
+  vtkDistancePolyDataFilter *PolyDataDistance;
 
 };
 

### 2011-05-20 Source/vtkDistancePolyDataFilter.h

@@ -1,7 +1,7 @@
 /*=========================================================================
 
   Program:   Visualization Toolkit
-  Module:    vtkPolyDataDistance.h
+  Module:    vtkDistancePolyDataFilter.h
 
   Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
   All rights reserved.
@@ -12,7 +12,7 @@
      PURPOSE.  See the above copyright notice for more information.
 
 =========================================================================*/
-// .NAME vtkPolyDataDistance
+// .NAME vtkDistancePolyDataFilter
 // .SECTION Description
 //
 // Computes the signed distance from one vtkPolyData to another. The
@@ -29,15 +29,15 @@
 // Written by Chris Weigle and Cory Quammen, The University of North
 // Carolina at Chapel Hill.
 
-#ifndef __vtkPolyDataDistance_h
-#define __vtkPolyDataDistance_h
+#ifndef __vtkDistancePolyDataFilter_h
+#define __vtkDistancePolyDataFilter_h
 
 #include "vtkPolyDataAlgorithm.h"
 
-class vtkPolyDataDistance : public vtkPolyDataAlgorithm {
+class vtkDistancePolyDataFilter : public vtkPolyDataAlgorithm {
 public:
-  static vtkPolyDataDistance *New();
-  vtkTypeMacro(vtkPolyDataDistance, vtkPolyDataAlgorithm);
+  static vtkDistancePolyDataFilter *New();
+  vtkTypeMacro(vtkDistancePolyDataFilter, vtkPolyDataAlgorithm);
   void PrintSelf(ostream& os, vtkIndent indent);
 
   // Description:
@@ -67,8 +67,8 @@ public:
   vtkPolyData* GetSecondDistanceOutput();
 
 protected:
-  vtkPolyDataDistance();
-  ~vtkPolyDataDistance();
+  vtkDistancePolyDataFilter();
+  ~vtkDistancePolyDataFilter();
 
   int RequestData(vtkInformation*, vtkInformationVector**, vtkInformationVector*);
   int FillInputPortInformation(int, vtkInformation*);
@@ -80,8 +80,8 @@ protected:
   int ComputeSecondDistance;
 
 private:
-  vtkPolyDataDistance(const vtkPolyDataDistance&); // no implementation
-  void operator=(const vtkPolyDataDistance&);      // no implementation
+  vtkDistancePolyDataFilter(const vtkDistancePolyDataFilter&); // no implementation
[…]

### 2011-05-20 Source/vtkBooleanOperationPolyDataFilter.h
Renamed vtkPolyDataIntersection to vtkIntersectionPolyDataFilter.

@@ -35,7 +35,7 @@
 
 class vtkIdList;
 class vtkDistancePolyDataFilter;
-class vtkPolyDataIntersection;
+class vtkIntersectionPolyDataFilter;
 
 
 class vtkBooleanOperationPolyDataFilter : public vtkPolyDataAlgorithm
@@ -120,7 +120,7 @@ private:
                  vtkDataSetAttributes::FieldList & cellFieldList,
                  vtkIdList* cellIds, bool reverseCells);
 
-  vtkPolyDataIntersection *PolyDataIntersection;
+  vtkIntersectionPolyDataFilter *PolyDataIntersection;
 
   vtkDistancePolyDataFilter *PolyDataDistance;
 

### 2011-05-20 Source/vtkIntersectionPolyDataFilter.h

@@ -1,7 +1,7 @@
 /*=========================================================================
 
   Program:   Visualization Toolkit
-  Module:    vtkPolyDataIntersection.h
+  Module:    vtkIntersectionPolyDataFilter.h
 
   Copyright (c) Ken Martin, Will Schroeder, Bill Lorensen
   All rights reserved.
@@ -12,10 +12,10 @@
      PURPOSE.  See the above copyright notice for more information.
 
 =========================================================================*/
-// .NAME vtkPolyDataIntersection
+// .NAME vtkIntersectionPolyDataFilter
 // .SECTION Description
 //
-// vtkPolyDataIntersection computes the intersection between two
+// vtkIntersectionPolyDataFilter computes the intersection between two
 // vtkPolyData objects. The first output is a set of lines that marks
 // the intersection of the input vtkPolyData objects. The second and
 // third outputs are the first and second input vtkPolyData,
@@ -25,17 +25,17 @@
 // Written by Chris Weigle and Cory Quammen, The University of North
 // Carolina at Chapel Hill.
 
-#ifndef __vtkPolyDataIntersection_h
-#define __vtkPolyDataIntersection_h
+#ifndef __vtkIntersectionPolyDataFilter_h
+#define __vtkIntersectionPolyDataFilter_h
 
 #include "vtkPolyDataAlgorithm.h"
 
 
-class vtkPolyDataIntersection : public vtkPolyDataAlgorithm
+class vtkIntersectionPolyDataFilter : public vtkPolyDataAlgorithm
 {
 public:
-  static vtkPolyDataIntersection *New();
-  vtkTypeMacro(vtkPolyDataIntersection, vtkPolyDataAlgorithm);
+  static vtkIntersectionPolyDataFilter *New();
+  vtkTypeMacro(vtkIntersectionPolyDataFilter, vtkPolyDataAlgorithm);
   virtual void PrintSelf(ostream &os, vtkIndent indent);
 
   // Description:
@@ -63,8 +63,8 @@ public:
                                           int &coplanar, double pt1[3], double pt2[3]);
 
 protected:
-  vtkPolyDataIntersection();
-  ~vtkPolyDataIntersection();
+  vtkIntersectionPolyDataFilter();
+  ~vtkIntersectionPolyDataFilter();
 
   int RequestData(vtkInformation*, vtkInformationVector**, vtkInformationVector*);
   int FillInputPortInformation(int, vtkInformation*);
[…]
